---
title: "Abalone Age Prediction"
author: "Edward Sobczak"
format: html
---

## Title


## Summary


## Introduction



## Methods and Results

The first analysis step is to load necessary packages and set the working directory.

```{r}
setwd(getwd())
library(tidyverse)
```

The second analysis read in the dataset from the internet. In our case, the Abalone dataset from the UCI Machine Learning Repository, and is stored in a zipped folder. The following code reads in the data and downloads it to the `data` folder of the working directory. This code **only needs to be run once**, so has now been made display-only.

```r
read_data <- function(file_path) {
  temp <- tempfile()
  zipped <- download.file(url = file_path, destfile = temp)
  file <- unz(temp, "abalone.data")
  abalone_data <- read_csv(file, col_names = c("sex", 
                                               "length", 
                                               "diameter", 
                                               "height", 
                                               "whole_weight", 
                                               "shucked_weight", 
                                               "viscera_weight", 
                                               "shell_weight", 
                                               "rings"))
  
  unlink(temp)
  
  write_csv(abalone_data, file = "data/abalone_data.csv")
}

read_data("https://archive.ics.uci.edu/static/public/1/abalone.zip")
```
While the data has been stored in the repository, it has not yet been read into R. The following code reads the data into R.

```{r}
abalone_data <- read_csv("data/abalone_data.csv")
```



## Discussion



## References
